Embedding Intuitionistic into Classical Logic

### 2023-03-15 LPAR/LPAR-EIICL.tex
typo
small fixes

For predicate logic we assume an ambient signature $\Sigma$, i.e., a set of function and predicate symbols with their respective arities, and a set of variables. A term is either a variable or, given an $n$-ary function symbol $f$ and terms $t_1, \ldots, t_n$, $f(t_1, \ldots, t_n)$. A formula is either $\bot$, or, given a $n$-ary predicate symbol $P$ and terms $t_1, \ldots, t_n$, $P(t_1, \ldots, t_n)$, or, given formulae $\varphi$ and $\psi$ and $\circ\in\{\wedge, \vee,\to\}$, $\varphi\circ\psi$, or, given a formula $\varphi$ and a variable $x$, $\exists x\varphi$ or $\forall x\varphi$. Free and bound variables are defined as usual and there is a Barendregt's convention in place, i.e. we implicitly assume that they comprise distinct sets, such that bound variables are not affected by substitutions of free variables.

@@

\begin{definition}
	Let $\Sigma$ be a signature. A $\Sigma$-structure $\mathcal{M}$ consists of a non-empty set $M$, the \emph{domain} of $\mathcal{M}$, and an \emph{interpretation} $I$ that assigns
	\begin{itemize}
		\item to each $n$-ary function symbol $f\in\Sigma$ a $n$-ary function $f^I: M^n\to M$.
		\item to each $n$-ary predicate symbol $R\in\Sigma$ a $n$-ary predicate $R^I\subseteq M^n$.
	\end{itemize}
	A variable assignment $v$ is a function that assigns to each free variable an element $m\in M$. For each free variable $a$ and $m\in M$ we define $$v[m/a](b) = \begin{cases}
	m, &\text{if $b=a$,}\\
	v(b), &\text{otherwise.}
	\end{cases}$$
	Then terms are interpreted as follows:
	\begin{itemize}
		\item $a^{I, v} = v(a)$ for each free variable $a$.
		\item $f(t_1,\dots,t_n)^{I, v} = f^I(t_1^{I, v},\dots, t_n^{I, v})$ for each $n$-ary function symbol $f\in\Sigma$.
	\end{itemize}
	We define a model relation between pairs $\mathcal M, v$ and formulae $\varphi$ as follows
	\begin{itemize}
		\item $\mathcal M, v\not\models\bot$
		\item $\mathcal M, v\models R(t_1,\dots,t_n)$ iff $(t_1^{I, v},\dots,t_n^{I, v})\in R^I$, $\mathcal M, v\models s = t$ iff $s^{I, v} = t^{I, v}$.
		\item $\mathcal M, v\models \varphi\wedge \psi$ iff $\mathcal M, v\models\varphi$ and $\mathcal M, v\models\psi$
		\item $\mathcal M, v\models \varphi\vee\psi$ iff $\mathcal M, v\models\varphi$ or $\mathcal M, v\models\psi$.
		\item $\mathcal M, v\models \varphi\to\psi$ iff $\mathcal M, v\not\models\varphi$ or $\mathcal M, v\models\psi$.
		\item $\mathcal M, v\models\exists x\varphi$ iff there exists $m\in M$ such that $\mathcal M, v[m/a]\models\varphi[a/x]$.
		\item $\mathcal M, v\models\forall x\varphi$ iff for all $m\in M$ we have $\mathcal M, v[m/a]\models\varphi[a/x]$.
	\end{itemize}
	 Here $a$ is a free variable that does not occur in $\varphi$. $\varphi$ is satisfiable if $\mathcal M, v\models\varphi$ for some $\mathcal M, v$. It is valid if $\mathcal M, v\models\varphi$ for every $\mathcal M, v$. We denote the set of valid formulae with \emph{CQC} (Classical Quantified Calculus).
\end{definition}

@@

\begin{definition}
For $\psi\in\mathcal X$ let $[\psi]$ denote the corresponding clause in $\mathcal X^{b S}$. Suppose we are given a counter-model $\mathcal M = (M, I)$ for $\varphi^{CH}$ such that $m\preceq f_\psi^I(m)$ for each $m\in M$, $\psi\in \mathcal X$.
We say $(A\to B)\to C\in\mathcal X$ is \emph{fulfilled} at $m\in M$ iff $A^I(m)\to B^I(m)$ is false or $C^I(m)$ is true. For $\psi\in\mathcal X$ define $$g_\psi : M\to M, m\mapsto\begin{cases}
		m,&\text{ if $\psi$ is fulfilled at $m$ or $m\not\preceq f^I_{[\psi]}(m)$,}\\
		f^I_{\lbrack\psi\rbrack}(m),&\text{ else.}		
	\end{cases}$$Define a model $\mathcal M_T = (M_T, I_T)$ as follows:
	\begin{itemize}
		\item $M_T$ is the set of sequences without repetition on $\mathcal X$.
		\item $s^{I_T} = \epsilon$.
		\item Interpret $\preceq$ as the prefix-order.
		\item We set $$f_\psi^{I_T}(\psi_1\dots\psi_n) = \begin{cases}
			\psi_1\dots\psi_n, &\text{if $\psi$ occurs in $[\psi_1]\dots[\psi_n]$,}\\
			\psi_1\dots\psi_n\psi, &\text{else.}			
		\end{cases}$$
		\item For propositional variables $P$, we set $$P^{I_T}\left(\psi_1\dots \psi_n\right) = P^I\left(g_{\psi_n}\left(\dots\left(g_{\psi_1}\left(s^I\right)\right)\dots\right)\right).$$
	\end{itemize}
\end{definition}

@@

\begin{corollary}\label{cor:prop-tree-model}
	If $\mathcal M$ is a counter-model to $\varphi^{CH}$ then so is $\mathcal M_T$.
\end{corollary}

@@

	\emph{Case 1:} $\psi = \psi_0\to \psi_1$.\\
	Suppose first that $w_k, v_i\not\models \psi_0\to\psi_1$. Then there exists $w_l\geq w_k$ such that $w_l, v_i\models \psi_0$ and $w_l, v_i\not\models\psi_1$. Then by induction hypothesis $\mathcal M, v_c[w_l/b]\models \psi_0^b$ and $\mathcal M, v_c[w_l/b]\models \psi_1^b$, i.e. $\mathcal M, v_c[w_l/b]\models \psi$. Suppose on the other hand that $w_l, v_i\models \psi_0\to\psi_1$. Then for all $w_k\geq w_l$ we have $w_k\models \psi_0\to \psi_1$, i.e. $w_k, v_i\not\models\psi_0$ or $w_k, v_i\models\psi_1$. Then by induction hypothesis $\mathcal M, v_c[w_k/b]\not\models\psi_0$ or $\mathcal M, v_c[w_k/b]\models\psi_1$, i.e. $\mathcal M, v_c[w_k/b]\models \psi_0\to\psi_1$.

>>>>

For predicate logic we assume an ambient signature $\Sigma$, i.e., a set of function and predicate symbols with their respective arities, and a set of variables. A term is either a variable or, given an $n$-ary function symbol $f$ and terms $t_1, \ldots, t_n$, $f(t_1, \ldots, t_n)$. A formula is either $\bot$, or, given a $n$-ary predicate symbol $P$ and terms $t_1, \ldots, t_n$, $P(t_1, \ldots, t_n)$, or, given formulae $\varphi$ and $\psi$ and $\circ\in\{\wedge, \vee,\to\}$, $\varphi\circ\psi$, or, given a formula $\varphi$ and a variable $x$, $\exists x\varphi$ or $\forall x\varphi$. Free and bound variables are defined as usual and there is a Barendregt's convention in place, i.e. we implicitly assume that they comprise distinct sets, such that bound variables are not affected by substitutions of free variables in any formula.

@@

\begin{definition}
	Let $\Sigma$ be a signature. A $\Sigma$-structure $\mathcal{M}$ consists of a non-empty set $M$, the \emph{domain} of $\mathcal{M}$, and an \emph{interpretation} $I$ that assigns
	\begin{itemize}
		\item to each $n$-ary function symbol $f\in\Sigma$ a $n$-ary function $f^I: M^n\to M$.
		\item to each $n$-ary predicate symbol $R\in\Sigma$ a $n$-ary predicate $R^I\subseteq M^n$.
	\end{itemize}
	A variable assignment $v$ is a function that assigns to each free variable an element $m\in M$. For each free variable $a$ and $m\in M$ we define $$v[m/a](b) = \begin{cases}
	m, &\text{if $b=a$,}\\
	v(b), &\text{otherwise.}
	\end{cases}$$
	Then terms are interpreted as follows:
	\begin{itemize}
		\item $a^{I, v} = v(a)$ for each free variable $a$.
		\item $f(t_1,\dots,t_n)^{I, v} = f^I(t_1^{I, v},\dots, t_n^{I, v})$ for each $n$-ary function symbol $f\in\Sigma$.
	\end{itemize}
	We define a model relation between pairs $\mathcal M, v$ and formulae $\varphi$ as follows
	\begin{itemize}
		\item $\mathcal M, v\not\models\bot$
		\item $\mathcal M, v\models R(t_1,\dots,t_n)$ iff $(t_1^{I, v},\dots,t_n^{I, v})\in R^I$, $\mathcal M, v\models s = t$ iff $s^{I, v} = t^{I, v}$.
		\item $\mathcal M, v\models \varphi\wedge \psi$ iff $\mathcal M, v\models\varphi$ and $\mathcal M, v\models\psi$
		\item $\mathcal M, v\models \varphi\vee\psi$ iff $\mathcal M, v\models\varphi$ or $\mathcal M, v\models\psi$.
		\item $\mathcal M, v\models \varphi\to\psi$ iff $\mathcal M, v\not\models\varphi$ or $\mathcal M, v\models\psi$.
		\item $\mathcal M, v\models\exists x\varphi$ iff there exists $m\in M$ such that $\mathcal M, v[m/x]\models\varphi$.
		\item $\mathcal M, v\models\forall x\varphi$ iff for all $m\in M$ we have $\mathcal M, v[m/x]\models\varphi$.
	\end{itemize}
	$\varphi$ is satisfiable if $\mathcal M, v\models\varphi$ for some $\mathcal M, v$. It is valid if $\mathcal M, v\models\varphi$ for every $\mathcal M, v$. We denote the set of valid formulae with \emph{CQC} (Classical Quantified Calculus).
\end{definition}

@@

\begin{definition}
For $\psi\in\mathcal X$ let $[\psi]$ denote the corresponding clause in $\mathcal X^{b S}$. Suppose we are given a counter-model $\mathcal M = (M, I)$ for $\varphi^{CH}$.
We say $(A\to B)\to C\in\mathcal X$ is \emph{fulfilled} at $m\in M$ iff $A^I(m)\to B^I(m)$ is false or $C^I(m)$ is true. For $\psi\in\mathcal X$ define $$g_\psi : M\to M, m\mapsto\begin{cases}
		m,&\text{ if $\psi$ is fulfilled at $m$ or $m\not\preceq f^I_{[\psi]}(m)$,}\\
		f^I_{\lbrack\psi\rbrack}(m),&\text{ else.}		
	\end{cases}$$Define a model $\mathcal M_T = (M_T, I_T)$ as follows:
	\begin{itemize}
		\item $M_T$ is the set of sequences without repetition on $\mathcal X$.
		\item $s^{I_T} = \epsilon$.
		\item Interpret $\preceq$ as the prefix-order.
		\item We set $$f_\psi^{I_T}(\psi_1\dots\psi_n) = \begin{cases}
			\psi_1\dots\psi_n, &\text{if $\psi$ occurs in $[\psi_1]\dots[\psi_n]$,}\\
			\psi_1\dots\psi_n\psi, &\text{else.}			
		\end{cases}$$
		\item For propositional variables $P$, we set $$P^{I_T}\left(\psi_1\dots \psi_n\right) = P^I\left(g_{\psi_n}\left(\dots\left(g_{\psi_1}\left(s^I\right)\right)\dots\right)\right).$$
	\end{itemize}
\end{definition}

@@

@@

	\emph{Case 1:} $\psi = \psi_0\to \psi_1$.\\
	Suppose first that $w_k, v_i\not\models \psi_0\to\psi_1$. Then there exists $w_l\geq w_k$ such that $w_l, v_i\models \psi_0$ and $w_l, v_i\not\models\psi_1$. Then by induction hypothesis $\mathcal M, v_c[w_l/b]\models \psi_0^b$ and $\mathcal M, v_c[w_l/b]\not\models \psi_1^b$, i.e. $\mathcal M, v_c[w_l/b]\not\models \psi$. Suppose on the other hand that $w_l, v_i\models \psi_0\to\psi_1$. Then for all $w_k\geq w_l$ we have $w_k\models \psi_0\to \psi_1$, i.e. $w_k, v_i\not\models\psi_0$ or $w_k, v_i\models\psi_1$. Then by induction hypothesis $\mathcal M, v_c[w_k/b]\not\models\psi_0$ or $\mathcal M, v_c[w_k/b]\models\psi_1$, i.e. $\mathcal M, v_c[w_k/b]\models \psi_0\to\psi_1$.
